---
title: "Revision"
date: "`r Sys.Date()`"
output: html_document
---

```{r DoNotModify, include=FALSE}
### Utilities. Do not modify.
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos = "https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, FUN = InstallPackage))
}

# Basic packages
InstallPackages(c("bookdown", "formatR", "ragg", "magick", "kableExtra"))

# Chunk font size hook: allows size='small' or any valid Latex font size in chunk options
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(
    options$size != "normalsize", 
    paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), 
    x
  )
})
```

```{r Options, include=FALSE}
### Customized options for this document
# Add necessary packages here
Packages <- c(
  "tidyverse", 
  "spatstat", 
  "dbmss", 
  "pbapply", 
  "plyr", 
  "GoFKernel", 
  "microbenchmark", 
  "profmem",
  "gridExtra"
)
# Install them
InstallPackages(Packages)

# knitr options
knitr::opts_chunk$set(
  cache =   TRUE,     # Cache chunk results
  include = TRUE,     # Show/Hide chunks
  echo =    TRUE,     # Show/Hide code
  warning = FALSE,    # Show/Hide warnings
  message = FALSE,    # Show/Hide messages
  # Figure alignment and size
  fig.align = 'center', out.width = '80%',
  # Graphic devices (ragg_png is better than standard png)
  dev = c("ragg_png", "pdf"),
  # Code chunk format
  tidy = FALSE, tidy.opts = list(blank = FALSE, width.cutoff = 50),
  size = "scriptsize", knitr.graphics.auto_pdf = TRUE
  )
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```

Exécuter le fichier Appendix avant celui-ci pour définir les fonctions.

# Eclaircie de M

## Préalable

Exécution de M sur le jeu de points complet et simulation de 100 éclaircies pour obtenir une enveloppe de confiance de l'effet de l'éclaircie.
Ou bien, exécuter ce code:

```{r}
library("tidyverse")
library("dbmss")

# Parameters
par_case_ratio <- 1/20
# Expected number of clusters
par_kappa <- 20
# Cluster radius
par_scale <-  0.1
# Point size
par_size_gamma_shape <- 0.95
par_size_gamma_scale  <- 10
# Statistics
r <- c((0:9) / 100, (2:10) / 20)

# Draw Poisson ppp
X_csr <- function(
    points_nb,
    case_ratio = par_case_ratio,
    size_gamma_shape = par_size_gamma_shape,
    size_gamma_scale = par_size_gamma_scale) {
  points_nb %>% 
    runifpoint() %>% 
    as.wmppp() ->
    X
  cases_nb <- round(points_nb *  case_ratio)
  controls_nb <- points_nb - cases_nb
  c(rep("Control", controls_nb), rep("Case", cases_nb)) %>% 
    as.factor() ->
    X$marks$PointType
  rgamma(
    X$n, 
    shape = size_gamma_shape, 
    scale = size_gamma_scale
  ) %>% 
    ceiling() ->
    X$marks$PointWeight
  X
}

# Draw Matérn ppp
X_matern <- function(
    points_nb,
    case_ratio = par_case_ratio,
    kappa = par_kappa,
    scale = par_scale,
    size_gamma_shape = par_size_gamma_shape,
    size_gamma_scale = par_size_gamma_scale) {
  cases_nb <- round(points_nb *  case_ratio)
  controls_nb <- points_nb - cases_nb
  # CSR controls
  controls_nb %>% 
    runifpoint() %>% 
    superimpose(
      # Matern cases
      rMatClust(
        kappa = kappa, 
        scale = scale, 
        mu = cases_nb / kappa
      ) 
    ) %>% 
    as.wmppp() ->
    X
  # Update the number of cases
  cases_nb <- X$n - controls_nb
  c(rep("Control", controls_nb), rep("Case", cases_nb)) %>% 
    as.factor() ->
    X$marks$PointType
  rgamma(
    X$n, 
    shape = size_gamma_shape, 
    scale = size_gamma_scale
  ) %>% 
    ceiling() ->
    X$marks$PointWeight
  X
}

# Compute M
X_to_M <- function(X) {
  X %>% 
    Mhat(r = r, ReferenceType = "Case") %>% 
    pull("M")    
}
```

## Fonction de calcul de l'enveloppe

```{r}
Envelope_thinning <- function(X, r, FUN = Mhat, thinning = 1/10, nsim = 100) {
  # Repeat thinning (code from dbmss::MEnvelope)
  the_envelope <- spatstat.explore::envelope(
    X,
    fun = FUN,
    nsim = nsim,
    nrank = 1,
    r = r,
    ReferenceType = "Case",
    CheckArguments = FALSE,
    simulate = expression(rthin(X, P = thinning)),
    verbose = TRUE,
    savefuns = TRUE
  )
  attr(the_envelope, "einfo")$H0 <- "Thinned"
  # Calculate confidence intervals
  Envelope <- dbmss::FillEnvelope(the_envelope, Alpha = 0.05, Global = TRUE)
  # No edge effect correction
  attr(the_envelope, "einfo")$valname <- NULL
  # Return
  return(the_envelope)
}
```

## Eclaircie de 1E5 à 1E3

```{r}
# Number of points
par_points_comp_nb <- 1E5 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/100

# Draw the original point pattern
pp_original_Matern <- X_matern(par_points_comp_nb)
# Repeat thinning (code from dbmss::MEnvelope)
env_M_5_3 <- Envelope_thinning(
  pp_original_Matern,
  r = r,
  FUN = Mhat,
  thinning = thinning,
  nsim = 1000
)

autoplot(env_M_5_3)

# Sims contains simulated values (one line per simulation)
Sims <- as.data.frame(attr(env_M_5_3, "simfuns"))[, -1]
Bias <- abs(rowMeans(Sims, na.rm = TRUE) - env_M_5_3$obs)
Variance <- apply(Sims, MARGIN = 1, FUN = var, na.rm = TRUE)
RMSE <- sqrt(Bias^2 + Variance) / env_M_5_3$obs
```

Conclusion: la diminution du nombre de points fait exploser la variance.
La figure montre l'erreur relative en fonction de la distance.

```{r}
tibble(RMSE, r) %>% 
  ggplot(aes(x = r, y = RMSE)) +
  geom_line()
```

## Eclaircie de 1E6 à 1E5

```{r}
# Number of points
par_points_comp_nb <- 1E6 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/10

# Draw the original point pattern
pp_original_Matern <- X_matern(par_points_comp_nb)
# Repeat thinning (code from dbmss::MEnvelope)
env_M_6_5 <- Envelope_thinning(
  pp_original_Matern,
  r = r,
  FUN = Mhat,
  thinning = thinning,
  nsim = 100
)

autoplot(env_M_6_5)

# Sims contains simulated values (one line per simulation)
Sims <- as.data.frame(attr(env_M_6_5, "simfuns"))[, -1]
Bias <- abs(rowMeans(Sims, na.rm = TRUE) - env_M_6_5$obs)
Variance <- apply(Sims, MARGIN = 1, FUN = var, na.rm = TRUE)
RMSE <- sqrt(Bias^2 + Variance) / env_M_6_5$obs
```

L'erreur relative est petite.

```{r}
tibble(RMSE, r) %>% 
  ggplot(aes(x = r, y = RMSE)) +
  geom_line()
```

# Eclaircie de Kd

Même approche que pour M

## Eclaircie de 1E6 à 1E5

```{r}
# Number of points
par_points_comp_nb <- 1E6 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/10

# Draw the original point pattern
pp_original_Matern <- X_matern(par_points_comp_nb)
# Repeat thinning (code from dbmss::MEnvelope)
env_Kd_6_5 <- Envelope_thinning(
  pp_original_Matern,
  r = r,
  FUN = Kdhat,
  thinning = thinning,
  nsim = 100
)

autoplot(env_Kd_6_5)

# Sims contains simulated values (one line per simulation)
Sims <- as.data.frame(attr(env_Kd_6_5, "simfuns"))[, -1]
Bias <- abs(rowMeans(Sims, na.rm = TRUE) - env_Kd_6_5$obs)
Variance <- apply(Sims, MARGIN = 1, FUN = var, na.rm = TRUE)
RMSE <- sqrt(Bias^2 + Variance) / env_Kd_6_5$obs
```

# Approche de l'article

## Comparaison des courbes

Tirage d'un grand jeu de points et éclaircie.

```{r}
#| label: XList
# Number of points
par_points_comp_nb <- 1E5 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/10
# Number of simulations
simulations_n <- 20

# Simulate Matérn point patterns
X_matern_list <- replicate(
  simulations_n, 
  expr = X_matern(par_points_comp_nb), 
  simplify = FALSE
)
# Thin the point pattern
X_matern_thinned_list <- lapply(
  X_matern_list, 
  FUN = rthin, 
  P = thinning
)
# Simulate CSR point patterns
X_csr_list <- replicate(
  simulations_n, 
  expr = X_csr(par_points_comp_nb), 
  simplify = FALSE
)
# Group points and compute M
X_csr_thinned_list <- lapply(
  X_csr_list, 
  FUN = rthin, 
  P = thinning
)
```

The M function is computed on each point pattern.

```{r}
#| label: MCompare
library("pbapply")
# Compute M
M_matern_original <- pbsapply(X_matern_list, FUN = X_to_M)
M_matern_thinned <- pbsapply(X_matern_thinned_list, FUN = X_to_M)
M_csr_original <- pbsapply(X_csr_list, FUN = X_to_M)
M_csr_thinned <- pbsapply(X_csr_thinned_list, FUN = X_to_M)
```

The average values of M across simulations are plotted.

```{r}
#| label: MapproxFig
library("gridExtra")
# Figure: exact and thinned M(r)
Mapproxfig <- function(M_original, M_thinned, tag) {
  tibble(
    r,
    Exact = rowMeans(M_original), 
    Thinned =  rowMeans(M_thinned)
  ) %>% 
    pivot_longer(
      cols = !r,
      names_to = "M",
      values_to = "value"
    ) %>% 
    ggplot(aes(x = r, y = value, color = M, shape = M, linetype  = M)) +
    geom_line() +
    geom_point() +
    labs(x = "Distance", y = "M", tag = tag) +
    theme(
      legend.position = c(0.75, 0.8),
      legend.background = element_blank(),
      legend.text = element_text(size = 10)
    ) 
}

# Plot
grid.arrange(
  Mapproxfig(M_csr_original, M_csr_thinned, "a") + 
    theme(legend.position = "none"),
  Mapproxfig(M_matern_original, M_matern_thinned, "b"),
  ncol = 2
)
```

## Comparaison des p-values

The following code compares the p-values to reject the null hypothesis of random location of points in both CSR and Matérn point patterns at each distance.

```{r}
#| label: MCompareFig
# Function to calculate the departure of the thinned point pattern M p-value
# against H0 from that of the real point pattern
p_departure <- function(
    X_list,
    M_simulations = simulations_n,
    verbose = interactive()
  ) {
  # Thin the points
  X_thinned_list <- lapply(
    X_list, 
    FUN = rthin, 
    P = thinning
  )  
  # Prepare arrays to save the results
  quantiles_exact <-
    quantiles_thinned <-
    matrix(NA, nrow = length(r) - 1, ncol = simulations_n)
  for (pp_i in seq_len(simulations_n)) {
    # Iterate across the point patterns
    if (verbose) cat("Point pattern", pp_i, "of", simulations_n,"\n")
    # Calculate M with simulations
    envelope_exact <- MEnvelope(
      X_list[[pp_i]],
      r = r,
      NumberOfSimulations = M_simulations,
      ReferenceType = "Case",
      parallel = TRUE,
      verbose = FALSE,
    )
    # Individual simulations are in an attribute.
    # Extract them in a dataframe, delete the first row (distances)
    # Each column contains M(r) for a simulation
    sims_exact <- as.data.frame(attr(envelope_exact, "simfuns"))[, -1]
    # Quantiles of each M(r) in the simulations
    quantiles_exact[, pp_i] <- sapply(
      # Remove distance 0
      seq_along(r)[-1],
      FUN = function(r_i) {
        # Quantile function of the simulated M(r)
        ecdf(as.numeric(sims_exact[r_i, ]))(envelope_exact$obs[r_i])
      }
    )
    # M thinned, idem
    envelope_thinned <- MEnvelope(
      X_thinned_list[[pp_i]],
      r = r,
      NumberOfSimulations = M_simulations,
      ReferenceType = "Case",
      parallel = TRUE,
      verbose = FALSE,
    )
    sims_thinned <- as.data.frame(attr(envelope_thinned, "simfuns"))[, -1]
    quantiles_thinned[, pp_i] <- sapply(
      # Remove distance 0
      seq_along(r)[-1],
      FUN = function(r_i) {
        # Quantile function of the simulated M(r)
        ecdf(as.numeric(sims_thinned[r_i, ]))(envelope_thinned$obs[r_i])
      }
    )
  }
  return(
    tibble(
      Exact = rowMeans(quantiles_exact),
      Thinned = rowMeans(quantiles_thinned),
    )
  )
}

# Parallelize M simulations 
# (https://ericmarcon.github.io/dbmss/articles/parallel.html)
library("doFuture")
plan(multisession)
departure_csr <- p_departure(X_csr_list)
departure_matern <- p_departure(X_matern_list)
plan(sequential)

# Plot the p-values
departure_csr %>% 
  mutate(Pattern = "CSR", Distance = r[-1]) %>% 
  bind_rows(
    departure_matern %>% 
      mutate(Pattern = "Matérn", Distance = r[-1])
  ) %>% 
  pivot_longer(Exact:Thinned) %>% 
  ggplot() +
  geom_line(aes(x = Distance, y = value, color = Pattern, linetype = name)) +
  geom_hline(yintercept = .975, linetype = 2) +
  labs(color = "Method", y = "p-value")
```

# Comparaison des temps de calcul

Remplacement du paragraphe Computing Time.

Paramétrage

```{r}
#| label: ParamsXSizeCode
X_sizes <- c(1000, 5000, 10000)
```

Fonctions:

```{r}
#| label: XtoxCode
# Compute M
X_to_M <- function(X) {
  X %>% 
    Mhat(r = r, ReferenceType = "Case") %>% 
    pull("M")
}
X_to_D <- function(X) {
  X %>% 
    Dhat(r = r, Cases = "Case") %>% 
    pull("iso")
}

X_to_Kd <- function(X) {
  X %>% 
    Kdhat(r = r, ReferenceType = "Case") %>% 
    pull("Kd")
}
```

Test :

```{r}
#| label: TestTime
library("microbenchmark")
test_time <- function(points_nb, FUN_DBM, times = 10L) {
  X <- X_csr(points_nb)
  microbenchmark(FUN_DBM(X), times = times) %>% 
    pull("time")
}

X_sizes %>% 
  sapply(FUN = test_time, FUN_DBM = X_to_M) %>% 
  as_tibble() %>%
  pivot_longer(cols = everything()) %>% 
  rename(V = name, M = value) %>% 
  bind_cols(
    X_sizes %>%
    sapply(FUN = test_time, FUN_DBM = X_to_D) %>% 
    as_tibble() %>%
    pivot_longer(cols = everything()) %>% 
    rename(V = name, D = value) %>% 
    select(-V)
  ) %>% 
  bind_cols(
    X_sizes %>%
    sapply(FUN = test_time, FUN_DBM = X_to_Kd) %>% 
    as_tibble() %>%
    pivot_longer(cols = everything()) %>% 
    rename(V = name, Kd = value) %>% 
    select(-V)
  ) %>%
  pivot_longer(cols = !V)  %>% 
  group_by(V, name) %>% 
  summarise(Time = mean(value) / 1E9, sd = sd(value) / 1E9) %>%
  rename(Function = name) %>% 
  inner_join(
    tibble(
      V = paste0("V", seq_along(X_sizes)), 
      Size = X_sizes
    )
  ) -> dbm_time
```

Figure :

```{r}
dbm_time %>% 
  ggplot(aes(x = Size, y = Time, color = Function)) +
    geom_point() +
    geom_errorbar(aes(ymin = Time - sd, ymax = Time + sd)) +
    scale_x_log10() +
    scale_y_log10() + 
    labs(tag = "a") +
    theme(axis.text.x = element_text(size = 8))
```

```{r}
#| label: TimeModelCode
# Model
dbm_time %>% 
  filter(Function == "M") %>% 
  mutate(logTime = log(Time), logSize = log(Size)) ->
  M_time_log
M_time_lm <- lm(logTime ~ logSize, data = M_time_log) 
summary(M_time_lm)
```

**Présentation de la fonction: différence entre K, utilisée par Sweeney & Feser. Topographique.**

The computing time of the D function is actually around 100 longer than that of the M function because topographic DBMs require correcting edge effects that appear when the neighborhood of a point, i.e. the circle of radius $r$ around it, leaves the window. These correction are extremely computer intensive compared to simply counting the number of number of neighbors of each point or summing their weight.
Relative DBM's such as M do not face edge effects. The D function is thus not a realistic alternative for very large datasets

# Discussion

1.  D n'est pas une alternative parce que son temps de calcul est 10 à 100 fois supérieur à celui de M. Explication : D nécessite deux calculs de K et surtout des corrections d'effet de bord intensives en calcul. Accessoirement, le calcul de D sur 100000 points génère des pb de mémoire. La figure qui suit l'approche de l'article n'est pas intéressante.
2.  M comme D est stable à l'éclaircie : la moyenne des estimations de M après éclaircie égale exactement la vraie valeur de M (tirets rouges surperposés à la ligne noire dans tous les exemples).
3.  Si l'éclaircie est trop brutale (1000 points sur 100000), la variance de l'estimation explose, avec une erreur relative moyenne facilement supérieure à 100%. La technique est très efficace en revanche pour passer d'un million à 100000 points.
4.  Kd se comporte globalement comme M. Je pensais que ses calculs seraient beaucoup plus rapides mais ce n'est pas le cas. On n'en parle donc pas.
