---
title: "Revision"
date: "`r Sys.Date()`"
output: html_document
---

```{r DoNotModify, include=FALSE}
### Utilities. Do not modify.
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos = "https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, FUN = InstallPackage))
}

# Basic packages
InstallPackages(c("bookdown", "formatR", "ragg", "magick", "kableExtra"))

# Chunk font size hook: allows size='small' or any valid Latex font size in chunk options
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(
    options$size != "normalsize", 
    paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), 
    x
  )
})
```

```{r Options, include=FALSE}
### Customized options for this document
# Add necessary packages here
Packages <- c(
  "tidyverse", 
  "spatstat", 
  "dbmss", 
  "pbapply", 
  "plyr", 
  "GoFKernel", 
  "microbenchmark", 
  "profmem",
  "gridExtra"
)
# Install them
InstallPackages(Packages)

# knitr options
knitr::opts_chunk$set(
  cache =   TRUE,     # Cache chunk results
  include = TRUE,     # Show/Hide chunks
  echo =    TRUE,     # Show/Hide code
  warning = FALSE,    # Show/Hide warnings
  message = FALSE,    # Show/Hide messages
  # Figure alignment and size
  fig.align = 'center', out.width = '80%',
  # Graphic devices (ragg_png is better than standard png)
  dev = c("ragg_png", "pdf"),
  # Code chunk format
  tidy = FALSE, tidy.opts = list(blank = FALSE, width.cutoff = 50),
  size = "scriptsize", knitr.graphics.auto_pdf = TRUE
  )
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```

Exécuter le fichier Appendix avant celui-ci pour définir les fonctions.

# Eclaircie de M

## Préalable

Exécution de M sur le jeu de points complet et simulation de 100 éclaircies pour obtenir une enveloppe de confiance de l'effet de l'éclaircie.
Ou bien, exécuter ce code:

```{r}
library("tidyverse")
library("dbmss")

# Parameters
par_case_ratio <- 1/20
# Expected number of clusters
par_kappa <- 20
# Cluster radius
par_scale <-  0.1
# Point size
par_size_gamma_shape <- 0.95
par_size_gamma_scale  <- 10
# Statistics
r <- c((0:9) / 100, (2:10) / 20)

# Draw Matérn ppp
X_matern <- function(
    points_nb,
    case_ratio = par_case_ratio,
    kappa = par_kappa,
    scale = par_scale,
    size_gamma_shape = par_size_gamma_shape,
    size_gamma_scale = par_size_gamma_scale) {
  cases_nb <- round(points_nb *  case_ratio)
  controls_nb <- points_nb - cases_nb
  # CSR controls
  controls_nb %>% 
    runifpoint() %>% 
    superimpose(
      # Matern cases
      rMatClust(
        kappa = kappa, 
        scale = scale, 
        mu = cases_nb / kappa
      ) 
    ) %>% 
    as.wmppp() ->
    X
  # Update the number of cases
  cases_nb <- X$n - controls_nb
  c(rep("Control", controls_nb), rep("Case", cases_nb)) %>% 
    as.factor() ->
    X$marks$PointType
  rgamma(
    X$n, 
    shape = size_gamma_shape, 
    scale = size_gamma_scale
  ) %>% 
    ceiling() ->
    X$marks$PointWeight
  X
}

# Compute M
X_to_M <- function(X) {
  X %>% 
    Mhat(r = r, ReferenceType = "Case") %>% 
    pull("M")    
}
```

## Fonction de calcul de l'enveloppe

```{r}
Envelope_thinning <- function(X, r, FUN = Mhat, thinning = 1/10, nsim = 100) {
  # Repeat thinning (code from dbmss::MEnvelope)
  the_envelope <- spatstat.explore::envelope(
    X,
    fun = FUN,
    nsim = nsim,
    nrank = 1,
    r = r,
    ReferenceType = "Case",
    CheckArguments = FALSE,
    simulate = expression(rthin(X, P = thinning)),
    verbose = TRUE,
    savefuns = TRUE
  )
  attr(the_envelope, "einfo")$H0 <- "Thinned"
  # Calculate confidence intervals
  Envelope <- dbmss::FillEnvelope(the_envelope, Alpha = 0.05, Global = TRUE)
  # No edge effect correction
  attr(the_envelope, "einfo")$valname <- NULL
  # Return
  return(the_envelope)
}
```

## Eclaircie de 1E5 à 1E3

```{r}
# Number of points
par_points_comp_nb <- 1E5 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/100

# Draw the original point pattern
pp_original_Matern <- X_matern(par_points_comp_nb)
# Repeat thinning (code from dbmss::MEnvelope)
env_M_5_3 <- Envelope_thinning(
  pp_original_Matern,
  r = r,
  FUN = Mhat,
  thinning = thinning,
  nsim = 1000
)

autoplot(env_M_5_3)

# Sims contains simulated values (one line per simulation)
Sims <- as.data.frame(attr(env_M_5_3, "simfuns"))[, -1]
Bias <- abs(rowMeans(Sims, na.rm = TRUE) - env_M_5_3$obs)
Variance <- apply(Sims, MARGIN = 1, FUN = var, na.rm = TRUE)
RMSE <- sqrt(Bias^2 + Variance) / env_M_5_3$obs
```

Conclusion: la diminution du nombre de points fait exploser la variance.

## Eclaircie de 1E6 à 1E5

```{r}
# Number of points
par_points_comp_nb <- 1E6 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/10

# Draw the original point pattern
pp_original_Matern <- X_matern(par_points_comp_nb)
# Repeat thinning (code from dbmss::MEnvelope)
env_M_6_5 <- Envelope_thinning(
  pp_original_Matern,
  r = r,
  FUN = Mhat,
  thinning = thinning,
  nsim = 100
)

autoplot(env_M_6_5)

# Sims contains simulated values (one line per simulation)
Sims <- as.data.frame(attr(env_M_6_5, "simfuns"))[, -1]
Bias <- abs(rowMeans(Sims, na.rm = TRUE) - env_M_6_5$obs)
Variance <- apply(Sims, MARGIN = 1, FUN = var, na.rm = TRUE)
RMSE <- sqrt(Bias^2 + Variance) / env_M_6_5$obs
```




# Eclaircie de Kd

Même approche que pour M

## Eclaircie de 1E6 à 1E5

```{r}
# Number of points
par_points_comp_nb <- 1E6 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/10

# Draw the original point pattern
pp_original_Matern <- X_matern(par_points_comp_nb)
# Repeat thinning (code from dbmss::MEnvelope)
env_Kd_6_5 <- Envelope_thinning(
  pp_original_Matern,
  r = r,
  FUN = Kdhat,
  thinning = thinning,
  nsim = 100
)

autoplot(env_Kd_6_5)

# Sims contains simulated values (one line per simulation)
Sims <- as.data.frame(attr(env_Kd_6_5, "simfuns"))[, -1]
Bias <- abs(rowMeans(Sims, na.rm = TRUE) - env_Kd_6_5$obs)
Variance <- apply(Sims, MARGIN = 1, FUN = var, na.rm = TRUE)
RMSE <- sqrt(Bias^2 + Variance) / env_Kd_6_5$obs
```


# Approche de l'article

## Comparaison des courbes

Tirage d'un grand jeu de  points et éclaircie.

```{r}
#| label: XList
# Number of points
par_points_comp_nb <- 1E5 # 1E5 in the article
# Ratio of points to keep during thinning
thinning <- 1/10

# Simulate Matérn point patterns
X_matern_list <- replicate(
  simulations_n, 
  expr = X_matern(par_points_comp_nb), 
  simplify = FALSE
)
# Thin the point pattern
X_matern_thinned_list <- lapply(
  X_matern_list, 
  FUN = rthin, 
  P = thinning
)
# Simulate CSR point patterns
X_csr_list <- replicate(
  simulations_n, 
  expr = X_csr(par_points_comp_nb), 
  simplify = FALSE
)
# Group points and compute M
X_csr_thinned_list <- lapply(
  X_csr_list, 
  FUN = rthin, 
  P = thinning
)
```

The M function is computed on each point pattern.

```{r}
#| label: MCompare
library("pbapply")
# Compute M
M_matern_original <- pbsapply(X_matern_list, FUN = X_to_M)
M_matern_thinned <- pbsapply(X_matern_thinned_list, FUN = X_to_M)
M_csr_original <- pbsapply(X_csr_list, FUN = X_to_M)
M_csr_thinned <- pbsapply(X_csr_thinned_list, FUN = X_to_M)
```

The average values of M across simulations are plotted.

```{r}
#| label: MapproxFig
library("gridExtra")
# Figure: exact and thinned M(r)
Mapproxfig <- function(M_original, M_thinned, tag) {
  tibble(
    r,
    Exact = rowMeans(M_original), 
    Thinned =  rowMeans(M_thinned)
  ) %>% 
    pivot_longer(
      cols = !r,
      names_to = "M",
      values_to = "value"
    ) %>% 
    ggplot(aes(x = r, y = value, color = M, shape = M, linetype  = M)) +
    geom_line() +
    geom_point() +
    labs(x = "Distance", y = "M", tag = tag) +
    theme(
      legend.position = c(0.75, 0.8),
      legend.background = element_blank(),
      legend.text = element_text(size = 10)
    ) 
}

# Plot
grid.arrange(
  Mapproxfig(M_csr_original, M_csr_thinned, "a") + 
    theme(legend.position = "none"),
  Mapproxfig(M_matern_original, M_matern_thinned, "b"),
  ncol = 2
)
```

## Comparaison des p-values

The following code compares the p-values to reject the null hypothesis of random location of points in both CSR and Matérn point patterns at each distance.

```{r}
#| label: MCompareFig

# Function to calculate the departure of the thinned point pattern M p-value
# against H0 from that of the real point pattern
p_departure <- function(
    X_list,
    M_simulations = simulations_n,
    verbose = interactive()
  ) {
  # Thin the points
  X_thinned_list <- lapply(
    X_list, 
    FUN = rthin, 
    P = thinning
  )  
  # Prepare arrays to save the results
  quantiles_exact <-
    quantiles_thinned <-
    matrix(NA, nrow = length(r) - 1, ncol = simulations_n)
  for (pp_i in seq_len(simulations_n)) {
    # Iterate across the point patterns
    if (verbose) cat("Point pattern", pp_i, "of", simulations_n,"\n")
    # Calculate M with simulations
    envelope_exact <- MEnvelope(
      X_list[[pp_i]],
      r = r,
      NumberOfSimulations = M_simulations,
      ReferenceType = "Case",
      parallel = TRUE,
      verbose = FALSE,
    )
    # Individual simulations are in an attribute.
    # Extract them in a dataframe, delete the first row (distances)
    # Each column contains M(r) for a simulation
    sims_exact <- as.data.frame(attr(envelope_exact, "simfuns"))[, -1]
    # Quantiles of each M(r) in the simulations
    quantiles_exact[, pp_i] <- sapply(
      # Remove distance 0
      seq_along(r)[-1],
      FUN = function(r_i) {
        # Quantile function of the simulated M(r)
        ecdf(as.numeric(sims_exact[r_i, ]))(envelope_exact$obs[r_i])
      }
    )
    # M thinned, idem
    envelope_thinned <- MEnvelope(
      X_thinned_list[[pp_i]],
      r = r,
      NumberOfSimulations = M_simulations,
      ReferenceType = "Case",
      parallel = TRUE,
      verbose = FALSE,
    )
    sims_thinned <- as.data.frame(attr(envelope_thinned, "simfuns"))[, -1]
    quantiles_thinned[, pp_i] <- sapply(
      # Remove distance 0
      seq_along(r)[-1],
      FUN = function(r_i) {
        # Quantile function of the simulated M(r)
        ecdf(as.numeric(sims_thinned[r_i, ]))(envelope_thinned$obs[r_i])
      }
    )
  }
  return(
    tibble(
      Exact = rowMeans(quantiles_exact),
      Thinned = rowMeans(quantiles_thinned),
    )
  )
}

# Parallelize M simulations 
# (https://ericmarcon.github.io/dbmss/articles/parallel.html)
library("doFuture")
plan(multisession)
departure_csr <- p_departure(X_csr_list)
departure_matern <- p_departure(X_matern_list)
plan(sequential)

# Plot the p-values
departure_csr %>% 
  mutate(Pattern = "CSR", Distance = r[-1]) %>% 
  bind_rows(
    departure_matern %>% 
      mutate(Pattern = "Matérn", Distance = r[-1])
  ) %>% 
  pivot_longer(Exact:Thinned) %>% 
  ggplot() +
  geom_line(aes(x = Distance, y = value, color = Pattern, linetype = name)) +
  geom_hline(yintercept = .975, linetype = 2) +
  labs(color = "Method", y = "p-value")
```


# D

Question : comparer ou pas les temps de calcul de M, Kd et M.

```{r}
# Number of points
par_points_comp_nb <- 1E5 # 1E5 in the article
par_points_comp_nb %>% 
  X_matern() %>% 
  Dhat(Cases = "Case") %>% 
  autoplot()
  
```

