---
title: "Revision"
date: "`r Sys.Date()`"
output: html_document
---

# A user's guide to deal with large datasets

Computation time becomes critical when very large datasets are considered, i.e. above 100,000 points as shown in the previous section. The M function is often used to test a point pattern against a null hypothesis, requiring simulations of a null point pattern and the same computing time for each of them.
We explore here three possible ways to reduce the computational burden: using the D function insteaf of M, approximating the position of points, and thinning the point pattern.

## The D function as an alternative

**Présentation de la fonction: différence entre K, utilisée par Sweeney & Feser. Topographique.**

The computing time of the D function is actually around 100 longer than that of the M function because topographic DBMs require correcting edge effects that appear when the neighborhood of a point, i.e. the circle of radius $r$ around it, leaves the window. These correction are extremely computer intensive compared to simply counting the number of number of neighbors of each point or summing their weight.
Relative DBM's such as M do not face edge effects. The D function is thus not a realistic alternative for very large datasets

## Approximating the position of points

```{r}
#| label: ParamsSimsCode
simulations_n <- 20
```

### Two main effects

Unambiguously, approximating the position of the points generates a loss of information: in each grid cell, the distance between all the points is set to zero, and the distance between two points in different cells is approximated using the distance between the centroids of the two cells.
The first consequence is that **no spatial structure in each cell can be detected because the information is completely lost.** The merits of any distance-based methods are detecting the exact geographic scale(s) where any patterns of aggregation, dispersion or independence of points exist.
Grouping points at the centroid of cells erases any spatial structure under the size grid.
The same limit under the size grid is faced for distributions with multiple patterns of points (that is a coexistence of attraction and dispersion of points depending on the distance considered).
@Tidu2023 gathered an average of 320 points in every 377 municipalities of Sardinia.
However, the local structure of these points was overlooked.
Moreover, we suspect **bias in *M* estimation based on a scale of the order of the magnitude of the size of the cells**, which should decrease with distance, when the relative size of the cells becomes negligible.

### Test setting

The effect of the location approximation is tested on a set of aggregated points, similar to the real @Tidu2023 data, and on a completely random point pattern.
Both comprise 100,000 points such that groups include 250 points (100,000 points divided by 400 groups) on average.

```{r}
#| label: XList
# Number of points for comparison simulations
par_points_comp_nb <- 1E5 
# Simulate Matérn point patterns
X_matern_list <- replicate(
  simulations_n, 
  expr = X_matern(par_points_comp_nb), 
  simplify = FALSE
)
# Group points
X_matern_grouped_list <- lapply(
  X_matern_list, 
  FUN = group_points, 
  partitions = par_partitions
)
# Simulate CSR point patterns
X_csr_list <- replicate(
  simulations_n, 
  expr = X_csr(par_points_comp_nb), 
  simplify = FALSE
)
# Group points and compute M
X_csr_grouped_list <- lapply(
  X_csr_list, 
  FUN = group_points, 
  partitions = par_partitions
)
```

`r simulations_n` sets of completely random and of aggregated points (`r format(as.numeric(par_points_nb), scientific=FALSE, big.mark=",")` points with `r par_case_ratio * 100`% of Cases) were simulated.
The exact calculation and the calculation on the grid points were performed on each set of points to evaluate the effect of the approximation.

```{r}
#| label: MCompare
library("pbapply")
# Compute M
if (file.exists("data/MCompare.RData")) {
  # Do not compute if it has been saved before
  load("data/MCompare.RData")
} else {
  M_matern_original <- pbsapply(X_matern_list, FUN = X_to_M)
  M_matern_grouped <- pbsapply(X_matern_grouped_list, FUN = X_to_M)
  M_csr_original <- pbsapply(X_csr_list, FUN = X_to_M)
  M_csr_grouped <- pbsapply(X_csr_grouped_list, FUN = X_to_M)
  save(
    M_matern_original, M_matern_grouped, M_csr_original, M_csr_grouped,
    file = "data/MCompare.RData"
  )
}
```

(ref:Mapprox) Average estimate of *M* from the exact position of the points compared with the values obtained by grouping the points for CSR (a) and Matérn (b) point patterns.

```{r}
#| label: MapproxFig
#| include: true
#| fig.asp: 0.5
#| fig.cap: "(ref:Mapprox)"
#| fig.env: "figure*"
#| out.extra: ""

# Figure: exact and grouped M(r)
Mapproxfig <- function(M_original, M_grouped, tag) {
  tibble(
    r,
    Exact = rowMeans(M_original), 
    Grouped =  rowMeans(M_grouped)
  ) %>% 
    pivot_longer(
      cols = !r,
      names_to = "M",
      values_to = "value"
    ) %>% 
    ggplot(aes(x = r, y = value, color = M, shape = M, linetype  = M)) +
    geom_line() +
    geom_point() +
    labs(x = "Distance", y = "M", tag = tag) +
    theme(
      legend.position = c(0.75, 0.8),
      legend.background = element_blank(),
      legend.text = element_text(size = 10)
    ) 
}

# Plot
grid.arrange(
  Mapproxfig(M_csr_original, M_csr_grouped, "a") + 
    theme(legend.position = "none"),
  Mapproxfig(M_matern_original, M_matern_grouped, "b"),
  ncol = 2
)
```

### Results

The mean values of the estimates of *M* are presented in Figure \@ref(fig:MapproxFig).
The size of the grid cells is equal to `r 1/par_partitions`.
All neighbors at distances less than this threshold are placed at zero distance.
The estimate of the corresponding *M* function (Grouped *M*) is constant in that case, up to this threshold.
When the actual point pattern is not structured, some artifactual aggregation is generated at a small scale.
In contrast, the local aggregation of the Matérn pattern is underestimated.
Figure \@ref(fig:MapproxFig) shows that at substantially short distances the grouped *M* plot is below the exact *M* plot obtained using the actual position of points.

@Tidu2023 tested the effect of grouping the points by the correlation between exact and approximated *M* values.
We refrained from following them because the main source of variation of the grouped-point *M* values is that of the grouping itself: when the number of points per group is small, groups considerably vary between simulations, and the correlation is weak.
It increases with the size of the data (an illustration is given in the appendix).
A substantially high correlation does not exclude systematic errors.
Therefore it is not considered an appropriate statistic here.

(ref:MCompare) p-values to reject the null hypothesis (H0) of independence of the point locations tested by the *M* function. The colors of the curves represent CSR or aggregated point patterns. Solid lines provide the exact p-values of *M*, and dotted lines provide the values estimated on grouped point patterns. The horizontal, dotted line corresponds to the significance threshold, i.e., 97.5%, to reject H0 in a 5% risk-level two-tailed test.

```{r}
#| label: MCompareFig
#| include: true
#| fig.cap: "(ref:MCompare)"

# Function to calculate the departure of the grouped point pattern M p-value
# against H0 from that of the real point pattern
p_departure <- function(
    X_list,
    M_simulations = simulations_n,
    verbose = interactive()
  ) {
  # Group the points
  X_grouped_list <- lapply(
    X_list,
    FUN = group_points,
    partitions = par_partitions
  )
  # Prepare arrays to save the results
  quantiles_exact <-
    quantiles_grouped <-
    matrix(NA, nrow = length(r) - 1, ncol = simulations_n)
  for (pp_i in seq_len(simulations_n)) {
    # Iterate across the point patterns
    if (verbose) cat("Point pattern", pp_i, "of", simulations_n,"\n")
    # Calculate M with simulations
    envelope_exact <- MEnvelope(
      X_list[[pp_i]],
      r = r,
      NumberOfSimulations = M_simulations,
      ReferenceType = "Case",
      parallel = TRUE,
      verbose = FALSE,
    )
    # Individual simulations are in an attribute.
    # Extract them in a dataframe, delete the first row (distances)
    # Each column contains M(r) for a simulation
    sims_exact <- as.data.frame(attr(envelope_exact, "simfuns"))[, -1]
    # Quantiles of each M(r) in the simulations
    quantiles_exact[, pp_i] <- sapply(
      # Remove distance 0
      seq_along(r)[-1],
      FUN = function(r_i) {
        # Quantile function of the simulated M(r)
        ecdf(as.numeric(sims_exact[r_i, ]))(envelope_exact$obs[r_i])
      }
    )
    # M grouped, idem
    envelope_grouped <- MEnvelope(
      X_grouped_list[[pp_i]],
      r = r,
      NumberOfSimulations = M_simulations,
      ReferenceType = "Case",
      parallel = TRUE,
      verbose = FALSE,
    )
    sims_grouped <- as.data.frame(attr(envelope_grouped, "simfuns"))[, -1]
    quantiles_grouped[, pp_i] <- sapply(
      # Remove distance 0
      seq_along(r)[-1],
      FUN = function(r_i) {
        # Quantile function of the simulated M(r)
        ecdf(as.numeric(sims_grouped[r_i, ]))(envelope_grouped$obs[r_i])
      }
    )
  }
  return(
    tibble(
      Exact = rowMeans(quantiles_exact),
      Grouped = rowMeans(quantiles_grouped),
    )
  )
}

if (file.exists("data/MCompareFig.RData")) {
  # Do not compute if it has been saved before
  load("data/MCompareFig.RData")
} else {
  # Parallelize M simulations
  library("doFuture")
  plan(multisession)
  departure_csr <- p_departure(X_csr_list)
  departure_matern <- p_departure(X_matern_list)
  plan(sequential)
  save(
    departure_csr, departure_matern,
    file = "data/MCompareFig.RData"
  )
}
  
# Plot the p-values
departure_csr %>% 
  mutate(Pattern = "CSR", Distance = r[-1]) %>% 
  bind_rows(
    departure_matern %>% 
      mutate(Pattern = "Matérn", Distance = r[-1])
  ) %>% 
  pivot_longer(Exact:Grouped) %>% 
  ggplot() +
  geom_line(aes(x = Distance, y = value, color = Pattern, linetype = name)) +
  geom_hline(yintercept = .975, linetype = 2) +
  labs(color = "Method", y = "p-value")
```

We chose the mentioned test.
The *M* function is basically used as a test against the independence of point locations.
To assess the impact of grouping the points on the test result, Figure \@ref(fig:MCompareFig) shows its average p-value, computed among `r simulations_n` simulations of each point pattern.
At each distance, the *M* value is compared with that obtained with point patterns simulated based on the null hypothesis, which is rejected if the actual value is outside of the 95% central quantiles of the simulations.

In the case of CSR, the p-value to reject independence around 50%, regardless of the distance, and whether the points are grouped or not.
In the case of aggregated patterns, the null hypothesis is rejected correctly when the points are grouped.
At small scale, below the size of the cells, *M* values are actually that of the cell size: the test is correct because the point process we used is aggregated at all scales.
The point pattern might have been repulsive at the small scale, but this information is destroyed by grouping the points: *M* and their p-values are not reliable below the size of the cells.

**Since all information below the grid size is lost, the approximation might or might not be acceptable depending on the research question and the spatial scale at which interactions occur. Above the grid size, the *M* function is less affected by the approximation and the gap becomes rapidly negligible**.


## Random thinning

Random thinning consists of reducing the number of points by applying the same probability of being kept to each point. The resulting dataset is thus a representative sample of the actual one.

@Sweeney2005 showed that the K function, hence the D function, is invariant under random thinning. So is the M function: relative DBMs are obtained by dividing a number of neighbors of interest by a total number of neighbors. Both terms are reduced in the same proportion by random thinning, keeping the ratio unchanged.

Compared to approximating the position of points, random thinning does not degrade the information of each point but reduces the amount of data. The acceptability of the method can be tested rigorously by simulating randomly thinned point patterns and comparing the distribution of their M function to that of the original pattern. 

A million-point pattern is drawn from a Matérn process and thinned 1000 times. The M values of the original   and thinned patterns are compared.
Figure \@ref(fig:Mthin)a shows the confidence envelope of the thinned patterns. The relative Root Mean Squared Error (RMSE), that is the expectation of the error caused by thinning, is shown on Figure \@ref(fig:Mthin)b. It is very small, around 2%, at short distances and negligible at large distances.

**ici, les deux figures de "Eclaircie de 1E6 à 1E5" de Revision.Rmd, sous le nom fig:Mthin a et b**

As @Sweeney2005 already noted for the K function, the variance of the estimation of M increases when the number of points decreases. Thinning a 100,000-point dataset to 1000 points results in a very fast and unbiased estimation of the M function but RMSE is well over 100% at short distance (Figure in supplementary material). Thinning is an efficient method to reduce huge, untractable datasets to large, computable ones, but is not a solution to drastically reduce computing time by thinning large datasets to small ones.







